---
title: "Using tRophicPosition to estimate TP through the use of multiple models"
author: "Claudio Quezada-Romegialli, Andrew L Jackson & Chris Harrod"
date: "December 11 2022"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use of multiple models in tRophicPosition}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, eval = FALSE, echo = FALSE}
# To update sysdata
sysdata <- tRophicPosition:::sysdata
devtools::use_data(sysdata, internal = TRUE, overwrite = TRUE)
```

# tRophicPosition

`tRophicPosition`, is an R package incorporating a Bayesian model for the calculation of trophic position using stable isotopes with one or two baselines. As of 2022-12-11, the current version of the package is 0.8. `tRophicPosition` uses the powerful approach of Markov Chain Monte Carlo simulations provided by [JAGS](http://mcmc-jags.sourceforge.net) and the statistical language [R](https://www.r-project.org). Vignettes can be browsed with `browseVignettes("tRophicPosition")`.

# Introduction

In this vignette, we will introduce the new functions we have developed for `tRophicPosition` in the version 0.6.8. These functions are `multiModelTP()`, `credibilityIntervals()`, `pairwiseComparisons()`  and `parametricTP()`. Each of these functions accomplish different tasks within the Bayesian estimation of trophic position with stable isotopes, facilitating the calculation of multiple models (either one, two baselines and/or two baselines full model of trophic position) for a single species, pairwise comparisons of posterior estimates of trophic position and/or alpha parameters, plotting of credibility intervals of posterior estimates of trophic position and/or alpha, and calculation of a parametric (non-Bayesian) estimate of trophic position.

## Installing and loading the package

### Stable version - CRAN
First of all, you need to install [JAGS](http://mcmc-jags.sourceforge.net/) for your platform, and then install the stable version of `tRophicPosition` from [CRAN](https://CRAN.R-project.org/package=tRophicPosition):

```{r eval = FALSE}
install.packages("tRophicPosition")
```

After that, you have to load the package with:

```{r, eval = FALSE}
library(tRophicPosition)
```

### Development version - GitHub
If you want to install the development version of `tRophicPosition`, you must install it from GitHub. For this, we use the function `install_github()` from the package `devtools` (installation instructions [here](https://github.com/r-lib/devtools#updating-to-the-latest-version-of-devtools)), which needs to be installed previously (either from [CRAN](https://CRAN.R-project.org/package=devtools) or [GitHub](https://github.com/r-lib/devtools)):

```{r, eval = FALSE}
install.packages("devtools")
library(devtools)
```

If you are working in Windows, `devtools` also requires [Rtools](https://cran.r-project.org/bin/windows/Rtools/), or if you are working on a Mac, [Xcode](https://developer.apple.com/xcode/) (from Apple Store). In Linux you will need to install a compiler and various development libraries. 

Besides installing `devtools`, you must also install [JAGS](https://mcmc-jags.sourceforge.io/) (Just Another Gibbs Sampler), which is at the core of the Bayesian model analysis supporting `tRophicPosition`.

```{r, eval = FALSE}
install_github("clquezada/tRophicPosition", build_vignettes = TRUE)
```

After installing `tRophicPosition`, it should be loaded into memory, which automatically reports the version of the software (we need at least 0.6.8-8 to use the routines described in this vignette).

```{r}
library(tRophicPosition)
```

## Future releases and how to get support

You are encouraged to use `tRophicPosition` with your own data, test the package and see if there are any issues or problems. You can send your questions or commentaries to the google group [tRophicPosition-support](https://groups.google.com/d/forum/trophicposition-support) or directly to the email trophicposition-support@googlegroups.com. You can send your questions to https://stackexchange.com/ https://stackoverflow.com/ or even [Facebook (stable isotope ecology group)](https://www.facebook.com/groups/stableisotopes/).

We are constantly working on future releases of `tRophicPosition`, so feedback is very much appreciated.

# Calculation of TP for one species using multiple Bayesian models

To demonstrate how `tRophicPosition` works, we will simulate some data. For example, imagine we have a fish species of interest that has mean ($\pm$ SD) values of $\delta^{15}$N of 21&#8240; ($\pm$ 1&#8240;), and $\delta^{13}$C of -15&#8240; ($\pm$ 1&#8240;). Also, as in many aquatic ecosystems, this fish lives in a system with two sources of C and N, meaning that we have to deal with two different baselines. One of the baselines has mean ($\pm$ SD) values of $\delta^{15}$N of 14.2&#8240; ($\pm$ 1&#8240;) and $\delta^{13}$C -17.3&#8240; ($\pm$ 1&#8240;), while the other baseline has mean values of 15.4&#8240; ($\pm$ 1&#8240;) and -12.7&#8240; ($\pm$ 1&#8240;) ($\delta^{15}$N and $\delta^{13}$C respectively). We will use the function generateTPData() to simulate this simple ecosystem.

```{r}
consumer1 <- generateTPData(dCb1 = -17.3, dNb1 = 14.2,
                            dCc = -15, dNc = 21,
                            dCb2 = -12.7, dNb2 = 15.4,
                            DeltaN = 3.4, sd.DeltaN = 0.98,
                            DeltaC = 0.39, sd.DeltaC = 1.3,
                            consumer = "Consumer 1")
```

In this function we have also indicated mean and sd values for trophic discrimination factors for `DeltaN` ($\Delta$N) and `DeltaC` ($\Delta$C), and also have indicated a name for the consumer. Remember that all the functions have a help page where you will find other options (write for example `?generateTPData` in the console).

Now that we have our simple ecosystem simulated, we will plot it:

```{r, fig.width = 7, fig.height = 5}
plot(consumer1)
```

In this figure we can see that sample of simulated fish consumers is exactly located in-between both baselines on the $\delta^{13}$C axis, and also, our fish is 6.8&#8240; above the baseline 1 in terms of its $\delta^{15}$N value. If we use a one baseline model, and if the trophic level of baseline 1 (`lambda`) is 2, we would expect a mean trophic position of 4 for our simulated fish species. If we use a two baselines model without fractionation for carbon, we would expect an `alpha` value of 0.5, which means that each baseline is contributing in 50% of the energy inputs to our fish. Finally, if we use a two baselines full model (including fractionation for $\delta^{13}$C), the trophic position of our fish would depend on both baselines, and the relative contribution of each baseline would depend on the trophic discrimination factors ($\Delta$N and $\Delta$C). Clearly, this is worth exploring more, so we will calculate (and compare) estimates for TP using these models with the function `multiModelTP()`.

```{r, eval = FALSE}
consumer1_models <- multiModelTP(consumer1)
```

```{r, eval = FALSE, echo = FALSE}
# To update sysdata
consumer1_models <- multiModelTP(consumer1)
sysdata <- tRophicPosition:::sysdata
sysdata$vignetteMMCTP$consumer1_models <- consumer1_models

#Not sure why I put this here 
#Bilagay_models$consumer1_models$samples <- NULL
devtools::use_data(sysdata, internal = TRUE, overwrite = TRUE)
```

```{r, echo = FALSE}
consumer1_models <- multiModelTP(consumer1, n.adapt = 200, n.iter = 200, burnin = 200)
consumer1_models <- tRophicPosition:::sysdata$vignetteMMCTP$consumer1_models
```

`multiModelsTP()` requires an object of the class `isotopeData`, which was created previously by the function `generateTPData()`. By default `multiModelsTP()` defines a `lambda = 2` for the baselines, uses 2 chains (`n.chains = 2`) to do the Bayesian calculation with 20,000 adaptive iterations (`n.adapt = 20000`), 20,000 actual iterations (`n.iter = 20000`), 20,000 iterations as burnin (`burnin = 20000`) and a thinning of 10 (`thin = 10`). Furthermore, it estimates TP using three different Bayesian models: one baseline, two baselines and two baselines full (`models = c("oneBaseline", "twoBaselines", "twoBaselinesFull"`). The default setting is to limit the amount of output for each model run (print = FALSE). NB: The user can change any of these options, simply adding them as arguments to `multiModelTP()`, see `?multiModelTP()` for details.

Also, by default this function uses Post's (2002) assumptions on TDF (i.e. 56 values with mean 3.4 $\pm$ 0.98 SD for $\Delta$N and 107 values with mean 0.39 $\pm$ 1.3 SD for $\Delta$C).

For example, if we have used the option `print = TRUE` as argument (i.e. `multiModelTP(print = TRUE)`), we would have a trace plot for every model (to check graphically if our parameters of interest `TP`, `muDeltaN` and `alpha` have converged), and if more than 1 chain were used, also would have printed two statistics from the coda package: Gelman and Rubin's convergence diagnostic statistics (see ?gelman.diag for a detailed explanation). Basically, we expect that potential scale reduction factors for each parameter is close to 1, which means that our model has converged and it is reliable.

These estimates have been generated using Post’s (2002) mean estimates for TDFs, which are commonly used across a wide range of ecological studies. Now that we have calculated the trophic position for our simulated fish in a simple ecosystem, we will do the same calculation, but this time, we will change the trophic discrimination factor from Post's (2002) values to McCutchan's et al. (2003) values in order to examine the sensitivity of the models to TDF estimates. To do this, we will we will generate `consumer2` with the same isotope values as `consumer1` and simply change the variables `deltaN` and `deltaC` within the object `consumer2`.

```{r}
consumer2 <- generateTPData(dCb1 = -17.3, dNb1 = 14.2,
                            dCc = -15, dNc = 21,
                            dCb2 = -12.7, dNb2 = 15.4,
                            consumer = "Consumer 2")

consumer2$deltaN <- TDF(author = "McCutchan", element = "N")
consumer2$deltaC <- TDF(author = "McCutchan", element = "C")
```

And now we will calculate the three Bayesian models of trophic position again, with the only difference being we have changed their TDF values.

```{r, eval = FALSE}
consumer2_models <- multiModelTP(consumer2)
```

```{r, eval = FALSE, echo = FALSE}
# To update sysdata
consumer2_models <- multiModelTP(consumer2)
sysdata <- tRophicPosition:::sysdata
sysdata$vignetteMMCTP$consumer2_models <- consumer2_models

#Bilagay_models$consumer2_models$samples <- NULL #???
devtools::use_data(sysdata, internal = TRUE, overwrite = TRUE)
```

```{r, echo = FALSE}
consumer2_models <- multiModelTP(consumer2, n.adapt = 200, n.iter = 200, burnin = 200)
consumer2_models <- tRophicPosition:::sysdata$vignetteMMCTP$consumer2_models
```

In both cases we have calculated three Bayesian models (one baseline, two baselines without fractionation for C, and two baselines full model with C fractionation) for two simulated simple ecosystems consisting of one fish and 2 baselines, and the only difference between consumer 1 and consumer 2 is the trophic discrimination factor. In the first case we included Post's (2002) values, and in the second case we included McCutchan's et al. (2003) values for both N and C. We have saved the output of `multiModelTP()` into two objects `consumer1_models` and `consumer2_models`, and now we will analyze the results.

If you are using RStudio, in the environment tab you will see that both `consumer1_models` and `consumer2_models` are lists of 4 elements each. The first element is a data frame named TP. This dataframe has the posterior samples of TP for each model. The second element is also a data frame, named alpha in this case, which includes the posterior samples of alpha for 2 of the 3 models (only the two baselines models calculate alpha, i.e. the relative contribution of the baseline 1). The third element is a data frame as well, named gg (to indicate us that has the structure required to use the ggplot2 package). This data frame has summary values of both posterior samples of trophic position and alpha: median and 95% credibility interval, grouped by model, community and species. The fourth and last element is a list of 3 elements named samples, that has raw posterior mcmc samples if you want to summarise the information, or analyze it in a different way. The first three elements (data frames TP, alpha and gg) are based in these raw posterior samples of every Bayesian model, summarised by `multiModelsTP()` for convenience.

We can produce a detailed summary of the structure described above using the function `str()`:

``` {r, eval = FALSE}
str(consumer1_models)
```

In order to plot the mode and 95% credibility interval for both posterior trophic position and alpha, we can use the function `credibilityIntervals():`

```{r, fig.width = 6, fig.height = 4, warning=FALSE}
# For consumer 1 (based on Post's (2002) TDF values)
credibilityIntervals(consumer1_models$gg, x = "model")
```

This figure provides mode ($\pm$ 95% credibility interval) for posterior estimates trophic position from the three different models (one baseline two baselines and two baselines full), as well as alpha (only given for two baselines models).

```{r, fig.width = 6, fig.height = 4, warning=FALSE}
# For consumer 2 (based on McCutchan's (2003) TDF values)
credibilityIntervals(consumer2_models$gg, x = "model")
```

The arguments `df = consumer1_models$gg` and `x = model` above, tells the function `credibilityIntervals()` to plot the data frame `consumer1_models$gg` using the model column within that data frame as the independent variable (`x`), and by default using the mode as the central tendency descriptor (see `credibilityIntervals()` for details). If you want to plot the median instead of the mode add `y1 = "median` and `y2 = "alpha.median"` as arguments within `credibilityIntervals()`. When we have multiple species or several communities we will change this (e.g. see the vignette multiple species calculation of trophic position).

## Why we need of accurate trophic discrimination factors (TDF)?

Probably, the single and most important assumption to get accurate and reliable estimations of trophic position is the *trophic discrimination factor*, either for nitrogen $\Delta^{15}$N (one and two baselines models) and both $\Delta$N and $\Delta$C for the two baselines full model. Obviously, the use of a non-representative TDF estimate for either $\Delta$N or $\Delta$C can affect both trophic position estimations and the alpha parameter.

Now we will examine the impact of using the different TDFs for our simulated consumers. We have estimated trophic position and alpha for both consumers groups (differing only in TDF) and will test for statistical differences between posterior samples of these parameters. First, we must change the names of both posterior samples, because they have the same name (as they were calculated with `multiModelsTP()`):

```{r}
# Here we see that we have 4002 posterior samples of 3 parameters
# (one for each Bayesian model) for consumer1
str(consumer1_models$TP)

# And also 4002 posterior samples (for each 3 Bayesian models) for consumer2
str(consumer2_models$TP)

# But the names of each variables are the same for both consumers
# For consumer1
names(consumer1_models$TP)

# For consumer2
names(consumer2_models$TP)

# So, we change them in order to compare them.
# To make things clear, consumer 1 will be "Post" and consumer 2 will be
# "McCutchan". # Also, one baseline Bayesian model will be model1, two 
# baselines model will be model2 and two baselines full model will be model2F
names(consumer1_models$TP) <- c("Post-model1", 
                                "Post-model2",
                                "Post-model2F")
names(consumer2_models$TP) <- c("McCutchan-model1",
                                "McCutchan-model2",
                                "McCutchan-model2F")
```

A useful first step is to summarize the posterior distributions to get a quantitative idea of how different they are. For this, we will use the base function (i.e. that comes with R) `sapply()`. This function is part of the `lapply` functions that apply a function (in this case `summary`) over a list. As we have two lists (`consumer1_models$TP` and `consumer2_models$TP`) first we need to use the `c` operator, that combines them, then we will calculate a summary for the combined element:

```{r}
# Here we combine posterior estimates of trophic position for both consumers
combined_models <- c(consumer1_models$TP, consumer2_models$TP)

# Then we calculate a summary of posterior trophic position
sapply(combined_models, summary)

# And we calculate the modes
getPosteriorMode(combined_models)
```

Here we have a rudimentary summaries of posterior trophic position estimates associated with each model for both consumers. For example, the median TP for consumer 1 based on the one baseline model (Post-model1) is higher than consumer 1 calculated with both 2 baselines (Post-model2) and 2 baselines full (Post-model2F) models. It appears that the same is true for consumer 2, where median TP estimated for the one baseline model (McCutchan-model1) is higher than that for both two baselines model (McCutchan-model2 and McCutchan-model2F).

As we are operating in a Bayesian universe, there is little sense to state a threshold under or above which both distributions will be considered different (as used in a frequentist universe e.g. p > 0.05). Instead, we think rather of comparing posterior distributions. For example, we can say that two posterior trophic position estimates will be considered different if their samples of posterior distributions are 100% different. In this stringent case and considering Post-model1 and McCutchan-model1 posterior estimates of trophic position, we would expect then that all the posterior estimates of Post-model1 are lesser than posterior estimates of McCutchan-model1.

```{r}
compareTwoDistributions(combined_models$"Post-model1",
                        combined_models$"McCutchan-model1", 
                        test = "<=")
```

Above, we can see that trophic position posterior estimate of one baseline model using Post's (2002) TDF assumptions has a proportion of 4000 observations over the 4000 we sampled  (i.e. 4000 / 4000 = 1) that are lesser or equal than trophic position posterior estimate of the same one baseline model but using McCutchan's (2003) TDF assumptions. Thus, we are pretty confident that posterior estimation of trophic position calculated with one baseline Bayesian model with Post's (2000) assumptions is higher than posterior estimation of trophic position calculated with McCutchan's TDF assumptions. Remember that this is the same simulated fish consumer, in the same simple ecosystem, only considering differents TDF assumptions.

By default `compareTwoDistributions()` evaluates how many observations from `combined_models$"Post-model1"` are $\leq$ (less than or equal) than `combined_models$"McCutchan-model1"`, randomly drawn from each of the models posterior distribution. We can change to $\geq$, $<$ or $>$ stating the argument within `pairwiseComparisons` function (add `test = ">="`, for example). 

We can make a more detailed comparison of the different models and TDFs using the `pairwiseComparisons`()  function, which as its name suggests, produces a matrix of of pairwise comparisons:

```{r}
pairwiseComparisons(combined_models, test = "<=")
```

The matrix generated by this function is symmetrical, below the diagonal we have the results of quatifying the proportion of  posterior samples of the model in row are less or equal than posterior samples of the model in the column, while above the diagonal we have the results of quantifying the proportion of posterior samples of the model in the column are higher than posterior samples of the model in the row.

# Parametric trophic position
As an appendix, we will calculate the parametric (i.e. non-Bayesian) version of trophic position. For this, we have to iterate through each consumer, and do the calculation with `parametricTP()`.

```{r}
# First we combine both consumers' isotope values into a named list
consumers <- list("consumer1" = consumer1, "consumer2" = consumer2)

# And then, we calculate parametric TP using a loop for
for (consumer in consumers) parametricTP(consumer)
```
