---
title: "How to use tRophicPosition from scratch"
author: "Claudio Quezada-Romegialli & Chris Harrod"
date: "23th of May 2016"
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
  word_document: default
---

In this vignette, we will guide you to all the steps necessary to estimate the
trophic position of one or more species with stable isotopes, in a Bayesian framework.

# What is the trophic position

Briefly introduce the reader to the subject.

# Installing tRophicPosition

In order to use tRophicPosition, first you have to install it, either from GitHub, a package archive file or CRAN, and then load the library into memory.

## From GitHub

The easiest way to install the latest version of tRophicPosition is to install it from GitHub. In this way you always will have the code updated. You have to install the release version of devtools from CRAN, and **RTools** which is necessary to use it. Install RTools from here: http://cran.r-project.org/bin/windows/Rtools/. For other operating systems (Linux, Mac) refer to the [README](https://cran.r-project.org/web/packages/devtools/README.html) of devtools.

To install the devtools package from CRAN:
```{r eval = FALSE}
install.packages(devtools)
```

```{r eval = FALSE}
library(devtools)
install_github("AndrewLJackson/tRophicPosition")
```

## From a package archive file

## From CRAN

# Loading and saving data

## Loading data from the clipboard

## Loading data from a csv/txt file

## Saving the data

# Working with tRophicPosition

## Flow chart

First, we load the packages needed. Obviously [tRophicPosition](https://github.com/AndrewLJackson/tRophicPosition)
